---
output: github_document
bibliography: vignettes/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# asds2024.nils.practical

<!-- badges: start -->
<!-- badges: end -->

This module contains the final practical part for the Certificate of Advanced Studies
in Advanced Statistical Data Science ([CAS ASDS](https://www.unibe.ch/weiterbildungsangebote/cas_advanced_statistical_data_science/index_ger.html))
at the [University of Berne](https://www.unibe.ch/index_eng.html) for the class of 2024.

## Format

The practical module has been packaged as an R package, so everything (data, scripts, reports, ...)
should be contained in a single bundle, and analyses should be reproducible.

This setup follows suggestions from [@marwick2018-peerj; @marwick2018-tas], [@mflight2014], and [@rpackages2e]
(which provided more or less the instructions and toolchain recommendations based on which this package has been
created).

There are other opinions and tools (e.g. [@mflight2021] and [@R-targets; @targets2021]) for a lighter-weight
reproducible research approach, which I might explore in the future.

For more inspiration and available tools, see [@cran-view-rr].

## Installation

You can install the development version of `asds2024.nils.practical` from [GitHub](https://github.com/) with:

```r
# install.packages("devtools") # <- for `install_github` to be available uncomment this and run it (unless you've already installed it)

# Notes:
# - you probably want to install the suggested dependencies as well, since this package only uses suggested dependencies
# - when `install_github`-ing, you need to explicitly specify that you want the vignettes built as well
devtools::install_github(
  "nils-s/cas-asds-practical",
  dependencies = c("Depends", "Imports", "LinkingTo", "Suggests"),
  build_vignettes = TRUE)
```

Since not all documents are provided as vignettes, you probably want to clone the package sources into a local
directory as well:

```bash
git clone https://github.com/nils-s/cas-asds-practical.git
```

From there, you can more directly explore the raw data, and read documents that are not packaged as vignettes.

### Installation Troubleshooting

Assuming the `devtools` package is installed (and `install_github` is available), this package by itself should
not cause problems (simply because it contains very little stuff that could cause problems).
However, it depends on a bunch of dependencies, which will be installed when installing this package's suggested
dependencies as shown in the code snippet above.

The main suspect in this regard is the `sf` package, which has a few dependencies of its own (not all of which
are R packages). The first thing to try (after studying the error messages, of course) is to make sure all
prerequisites for `sf` are fulfilled (e.g. the [GEOS](https://libgeos.org), [GDAL](https://gdal.org),
and [PROJ](https://proj.org/) libraries).

On a Fedora machine, the following should get you started:

```bash
sudo dnf install gdal gdal-devel udunits2-devel proj proj-devel geos geos-devel
```

See [the `sf` documentation](https://r-spatial.github.io/sf/) for more information.

## Example

```{r}
library(asds2024.nils.practical)
vignette("get-started", package = "asds2024.nils.practical")
```

## References
