---
title: "Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes the data included in the package, how it was prepared, and why it was prepared in the way it was.
It does not describe the final form of the data as it is included in the package in detail (see the data description for
that, e.g. `?tracks`), but rather how the original raw data looked, and how and why it was pre-processed to end up the
way it is stored in the _data-raw_ folder (see package sources on [Github](https://github.com/nils-s/cas-asds-practical)).

```{r package-setup}
library(asds2024.nils.practical)
```

## Acquisition

The raw data was captured using a (now discontinued) bike computer of the model
["Sigma ROX GPS 11.0"](https://sigma.bike/product/rox-gps-11-0/), with an additional
chest strap heart rate sensor, and a combined speed and cadence sensor (both sensors
were bought bundled with the bike computer itself).

The data was exported using the (proprietary) Sigma software, and then exported as CSV files.

The data set comprises 157 trip recordings ("tracks"), each of which is a sequence
of data points taken at a five second interval. These data points were then aggregated
into one summarized row per track. Additionally, tracks were manually classified by
route, and direction in which the route was driven. These three data sets are available
in the package under the following names:

- `track_details` contains the recorded data point sequences of all tracks (approx. 160k rows)
- `tracks` contains the summarized track information, one row per track (157 rows)
- `track_classes` contains the classification information for each track, i.e. a class ID and direction (157 rows)

Tracks are identified by date, thus if, for example, summaries in `tracks` should be
extended with class labels for each track, `tracks` can be joined with `track_classes`
by the `date` key.

For convenience, there are two additional data sets included (`swiss_cantons` and
`swiss_lakes`), containing geographic data for the cantons and lakes of Switzerland.
These can be used to plot a geographic frame of reference when plotting tracks.
Since the trip routes are all within the canton of Berne, the canton data with a
`KTNR` ("Kantonsnummer") of 10 (i.e. Berne) is probably the most useful.
The source for the geo data is the Swiss Federal Statistical Office
(Bundesamt für Statistik, BfS), where the raw data can be downloaded from
[this page](https://www.bfs.admin.ch/bfs/de/home/dienstleistungen/geostat/geodaten-bundesstatistik/administrative-grenzen/generalisierte-gemeindegrenzen.assetdetail.30487000.html).

## Cleaning

The data available in the package is a cleaned-up version of the raw data available
in the package sources. Cleanups performed were:

- Some 0/0 GPS coordinates were set to `NA` values. These values were probably caused
  by some sort of GPS hiccup, so to clarify that there is no real data available for
  these points, and to prevent accidentally using the invalid coordinates for computations,
  they were set to `NA`.
- Some duplicate data points were removed. There were some data points in one
  track sequence that had the same _TrainingTimeAbsolute_ values. This looked like
  a temporary freeze of the bike computer, and there were some obviously-wrong
  values for distance and altitude changes in those rows, so only the first of
  those rows was kept, the others were completely removed.
- Power zone data was removed. The raw data contained columns for power zones,
  however, there is no power sensor installed or calibrated, so the raw data in
  these columns is bogus. To prevent confusion, this data was not included in the
  package data, i.e. power zone-related columns were removed.

The uncleaned data is still available in the raw CSV files available in the package
sources, and the _tracks.R_ file in the _data-raw_ folder contains the code that
was used to perform the cleanup.

## Geographic Censoring

Since the detailed track data points contain GPS data, the start and end points
of the tracks would show the exact starting points for the trips, i.e. my address.
Therefore, data points at the start and end of a trip, and within a certain region,
were removed. Additionally, the tracks were then adapted, so that the first point
of the censored track has a time and distance of 0, so that deducing the true
starting point of a track based on the censored data points is not possible
(or at least more difficult).

This approach of just cutting off parts of the track and resetting the first point
to 0 is slightly distorting the data, but this was deemed acceptable.
The initial idea of removing the first and last points few, but not resetting the
time and distance of the first uncensored point and instead adding random noise
to censor individual tracks but keeping the average track time and distance somewhat
more intact, was not pursued further because it is a more complex approach, and
the risk of accidentally leaking data was deemed higher.

The raw data for the tracks is only available in this censored form,
i.e. the CSV files in _data-raw/tracks/censored_ are the tracks with a trimmed start and end.
The original, unmodified raw CSV files are not publicly available.
The exact details of the censoring procedure can be found in the _data-raw/tracks.R_ file.

## Categorizing

The tracks are not all along different routes, some routes were driven multiple
times (some much more often than others). The tracks were (manually) classified
by their routes, with arbitrary numeric identifiers for the routes. Routes that
were only driven once did not get their own class, thus some tracks have a route
class of `NA`. Furthermore, one route was driven in different directions, tracks
along that route all have the same class, but are additionally labelled
as clockwise or counter-clockwise. All other routes were only ever driven in one
direction, these tracks are not labelled for direction (i.e. their direction column
contains `NA`).

Classification was done visually, by plotting the tracks, storing the plots as
image files, and then sorting them into folders. The commented code that was used
for classification can be found in _data-raw/track_classification.R_.
