---
title: "Praxismodul CAS ASDS 2024"
output: beamer_presentation
---

```{r notes_and_comments, include = FALSE}
# output options: beamer_presentation, slidy_presentation, ioslides_presentation

# ideas/TODOs:
# - random effects model, e.g. random effect for years
# - add cluster information to data (manually, or via one of the algorithms in the last part), model with that
# - (manually?) classify tracks
# - dynamic time warping
# - ?
```


```{r chunk_setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r package_setup, include = FALSE}
library(asds2024.nils.practical)
library(dplyr)
library(lubridate)
library(ggplot2)
library(sf)
library(ggspatial)
theme_set(theme_minimal())
library(broom)
library(glmnet)
library(cluster)
library(dbscan)
library(mclust)
```

# Data

---

```{r tracks_map, echo = FALSE}
be_fr <- swiss_cantons |>
  filter(KTNR %in% c(2,10)) # 2 = BE, 10 = FR
be_fr_lakes <- swiss_lakes |>
  filter(grepl("Biel|Brienz|Gruyère|Murten|Neuchâtel|Thun", GMDNAME)) # only show lakes named after these towns

track_details |>
  filter(!is.na(latitude) & !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = "WGS84") |> # EPSG 4326 = WGS-84
  st_transform(crs = 2056) |> # EPSG 2056 = CH-1903+/LV95
  to_linestrings() |>
  ggplot() +
  geom_sf(data = be_fr, fill = "#f0f0f0") +
  geom_sf(data = be_fr_lakes, fill = "#0080ff", alpha = 0.25) +
  geom_sf(color = "red", linewidth = 0.25, alpha = 0.2) +
  ggtitle("Geographical Visualization of Tracks") +
  annotation_scale(location = "bl", width_hint = 0.2) +
  annotation_north_arrow(
    location = "bl",
    pad_y = unit(1, "cm"),
    style = north_arrow_nautical,
    which_north = "true")
```

---

```{r tracks_per_year, echo = FALSE}
tracks |>
  group_by(year = year(date)) |>
  ggplot(aes(x = year)) +
  geom_bar() +
  labs(x = "Year", y = "Number of Tracks") +
  ggtitle("Number of Tracks per Year")
```

---

```{r tracks_per_month, echo = FALSE}
tracks |>
  group_by(month = month(date), year = year(date), .drop = FALSE) |>
  summarize(tracks = n(), .groups = "drop") |>
  ggplot(aes(x = month, y = tracks)) +
  geom_jitter(
    aes(
      fill = cut(year, 6, labels=as.character(2018:2023), include.lowest = TRUE),
      color = cut(year, 6, labels=as.character(2018:2023), include.lowest = TRUE),
      shape = cut(year, 6, labels=as.character(2018:2023), include.lowest = TRUE)
    ),
    width = 0.2,
    height = 0.15,
    alpha = 0.75,
    size = 5) +
  scale_shape_manual(values = c(21:24, 21, 25)) +
  labs(x = "Month", y = "Number of Tracks", fill = "Year", color = "Year", shape = "Year") +
  ggtitle("Monthly Tracks Count by Year")
```


# Analyses

---

## A First Linear Model

```{r basic_model, echo = FALSE, fig.show = "hold", out.width = "47.5%"}
fit_minimal <- lm(speed_km_h ~ distance_km + altitude_gain_m + avg_hr_bpm, data = tracks)
plot(fit_minimal)
```

---

```{r track2_map, echo = FALSE}
track_details |>
  filter(date == tracks[2,]$date) |>
  filter(!is.na(latitude) & !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = "WGS84") |> # EPSG 4326 = WGS-84
  st_transform(crs = 2056) |> # EPSG 2056 = CH-1903+/LV95
  ggplot(aes(color = distance_absolute_m)) +
  geom_sf(mapping = aes(color = distance_absolute_m), size = 0.5) +
  scale_color_gradient(low = "red", high = "green") +
  labs(color = "Distance (m)") +
  ggtitle("Track 2: Geographical Display")
```

---

```{r track2_distance, echo = FALSE}
track_details |>
  filter(date == tracks[2,]$date) |>
  ggplot(aes(x = training_time_absolute_s / 100, y = distance_absolute_m / 1000)) +
  geom_line() +
  labs(x = "Training Time (s)", y = "Distance (km)") +
  ggtitle("Track 2: Distance Over Time")
```

---

```{r track157_map, echo = FALSE}
track_details |>
  filter(date == tracks[157,]$date) |>
  filter(!is.na(latitude) & !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = "WGS84") |> # EPSG 4326 = WGS-84
  st_transform(crs = 2056) |> # EPSG 2056 = CH-1903+/LV95
  ggplot(aes(color = distance_absolute_m)) +
  geom_sf(mapping = aes(color = distance_absolute_m), size = 0.5) +
  # scale_color_gradient2(low = "blue", mid = "green", high = "red", midpoint = 40000)
  scale_color_gradientn(
    colors = c("red", "green", "blue", "blue"),
    values = scales::rescale(c(30, 42000, 45000, 114000), from = range(c(34, 114000)))) +
  labs(color = "Distance (m)") +
  ggtitle("Track 157: Geographical Display")
```

---

```{r track157_distance, echo = FALSE}
track_details |>
  filter(date == tracks[157,]$date) |>
  ggplot(aes(x = training_time_absolute_s / 100, y = distance_absolute_m / 1000)) +
  geom_line() +
  labs(x = "Training Time (s)", y = "Distance (km)") +
  ggtitle("Track 157: Distance Over Time")
```

---

```{r distance_correction, echo = FALSE}
corrected_track_details <- track_details |>
  mutate(
    correction = distance_absolute_m - lag(distance_absolute_m, default = 0) - distance_m,
    needs_correction = (distance_absolute_m > 90000) & (date == tracks[157,]$date),
    corrected_distance_absolute_m = ifelse(needs_correction, distance_absolute_m - max(correction), distance_absolute_m))

corrected_track_details |>
  filter(date == tracks[157,]$date) |>
  ggplot(aes(x = training_time_absolute_s / 100, y = corrected_distance_absolute_m / 1000)) +
  geom_line() +
  labs(x = "Training Time (s)", y = "Distance (km)") +
  ggtitle("Track 157: Corrected Distance Over Time")
```

---

```{r track157_corrected_map, echo = FALSE}
corrected_track_details |>
  filter(date == tracks[157,]$date) |>
  filter(!is.na(latitude) & !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = "WGS84") |> # EPSG 4326 = WGS-84
  st_transform(crs = 2056) |> # EPSG 2056 = CH-1903+/LV95
  ggplot(aes(color = distance_absolute_m)) +
  geom_sf(mapping = aes(color = corrected_distance_absolute_m), size = 0.5) +
  scale_color_gradient(low = "red", high = "green") +
  labs(color = "Distanz (m)") +
  ggtitle("Track 157 (corrected): Geographical Display")
```

---

## The Same Model, 2. Attempt

```{r basic_model_2nd_attempt, echo = FALSE, fig.show = "hold", out.width = "47.5%"}
corrected_tracks <- corrected_track_details |>
  mutate(
    distance_absolute_m = ifelse(needs_correction, corrected_distance_absolute_m, distance_absolute_m)
  ) |>
  summarize_tracks()

fit_corrected <- lm(speed_km_h ~ distance_km + altitude_gain_m + avg_hr_bpm, data = corrected_tracks)
plot(fit_corrected)
```

---

\begingroup
\fontsize{5}{10}

```{r basic_model_2_summary}
summary(fit_corrected)
```

\endgroup

## Temperature Effects

```{r temperature_model, echo = FALSE, fig.show = "hold", out.width = "47.5%"}
fit_temperature <- lm(speed_km_h ~ temperature_c, data = corrected_tracks)
plot(fit_temperature)
```

---

```{r temperature_data_vs_fitted, echo = FALSE}
plot_data_vs_fitted(fit_temperature, x = temperature_c, y = speed_km_h) +
  labs(x = "Temperature (°C)", y = "Speed (km/h)")
```

---

\begingroup
\fontsize{5}{10}

```{r temperature_model_summary}
summary(fit_temperature)
```

\endgroup

## Effect of Previous Training

```{r prev_training_model, echo = FALSE, fig.show = "hold", out.width = "47.5%"}
corrected_tracks <- corrected_tracks |>
  mutate(
    previous_training_date = lag(date),
    days_since_previous_training = as.numeric(difftime(date, previous_training_date, units = "days"))
  )
augmented_tracks <- corrected_tracks |>
  inner_join(corrected_tracks, join_by(previous_training_date == date), suffix = c("", "_previous_training"))
fit_pt <- lm(speed_km_h ~ distance_km + altitude_gain_m + avg_hr_bpm + days_since_previous_training + distance_km_previous_training + altitude_gain_m_previous_training, data = augmented_tracks)
plot(fit_pt)
```

---

\begingroup
\fontsize{5}{10}

```{r prev_training_model_summary}
summary(fit_pt)
```

\endgroup

## Training Effect over the Course of the Season

```{r month_model, echo = FALSE, warning = FALSE, fig.show = "hold", out.width = "47.5%"}
fit_month <- lm(speed_km_h ~ as.factor(month(date)), data = augmented_tracks)
plot(fit_month)
```

---

\begingroup
\fontsize{5}{10}

```{r month_model_summary}
summary(fit_month)
```

\endgroup

---

```{r distances_by_month, echo = FALSE}
corrected_tracks |>
  ggplot(aes(x = as.factor(month(date)), y = distance_km)) +
  geom_boxplot() +
  geom_jitter(width = 0.25, color = "#0000f0", alpha = 0.5) +
  labs(x = "Month", y = "Distance (km)") +
  ggtitle("Track Length by Month")
```

---

```{r month_interactions_model, echo = FALSE, warning = FALSE, fig.show = "hold", out.width = "47.5%"}
fit_month_and_more <- lm(speed_km_h ~ as.factor(month(date)) * (distance_km + avg_hr_bpm + altitude_gain_m), data = augmented_tracks)
# plot_data_vs_fitted(fit_month_and_more, x = `month(date)`, y = speed_km_h) +
#   labs(x = "Month", y = "Speed (km/h)")
plot(fit_month_and_more)
```

---

\begingroup
\fontsize{5}{10}

```{r month_interactions_model_summary}
summary(fit_month_and_more)
```

\endgroup

---

```{r calendar_week_model, echo = FALSE}
fit_calendar_week_and_more <- lm(speed_km_h ~ isoweek(date) * (distance_km + avg_hr_bpm + altitude_gain_m), data = augmented_tracks)
plot_data_vs_fitted(fit_calendar_week_and_more, x = `isoweek(date)`, y = speed_km_h) +
  labs(x = "Calendar Week", y = "Speed (km/h)")
```

---

\begingroup
\fontsize{5}{10}

```{r calendar_week_model_summary}
summary(fit_calendar_week_and_more)
```

\endgroup

## Elastic Net

```{r glmnet, echo = FALSE, fig.show = "hold", out.width = "47.5%"}
numerical_track_components <- corrected_tracks |>
  select(!date & !previous_training_date & !weekday & !samples)

training_set_ratio <- 0.7
training_set_indices <- 
  sample(1:nrow(numerical_track_components),
         nrow(numerical_track_components) * training_set_ratio) |>
  sort()
training_set <- numerical_track_components[training_set_indices,]
test_set <- numerical_track_components[setdiff(1:nrow(numerical_track_components), training_set_indices),]

# glmnet mixing param (hyperparameter; should be tuned instead of hard-coded)
alpha <- 0.9

fit_glm <- glmnet(
  x = training_set |> na.omit() |> select(!speed_km_h),
  y = training_set |> na.omit() |> select(speed_km_h) |> as.matrix(),
  alpha = alpha)
cv_glm <- cv.glmnet(
  x = training_set |> na.omit() |> select(!speed_km_h) |> as.matrix(),
  y = training_set |> na.omit() |> select(speed_km_h) |> as.matrix(),
  alpha = alpha)
plot(fit_glm, xvar = "lambda")
plot(cv_glm)
```

## PCA: Finding What's Important...

```{r pca_scree_plot, echo = FALSE}
numerical_track_components <- corrected_tracks |>
  mutate(weekday = as.numeric(weekday)) |>
  select(!date & !previous_training_date & !samples)
pca <- prcomp(~ ., data = numerical_track_components, scale = TRUE)
pca |>
  tidy("pcs") |> 
  ggplot(aes(x = PC, y = percent)) +
  geom_col() +
  # geom_point(aes(y = cumulative), color = "#8000ff") +
  labs(x = "PC", y = "Explained Variance")
```

---

```{r pca_biplot, echo = FALSE}
biplot(pca, scale = 0)
```

## Clustering

```{r cluster_data_prep, include = FALSE}
scaled_tracks <- corrected_tracks |>
  mutate(weekday = as.numeric(weekday)) |>
  select(!date & !previous_training_date & !samples) |>
  scale() |>
  as_tibble() |>
  na.omit()
```

```{r cluster_gaussian_mix, echo = FALSE}
clusters_gaussian_mix <- Mclust(scaled_tracks)
# plot(clusters_gaussian_mix, what = "classification")

scaled_tracks |>
  mutate(cluster = factor(clusters_gaussian_mix$classification)) |>
  ggplot(aes(x = distance_km, y = speed_km_h, color = cluster)) +
  geom_point() +
  labs(x = "Standardized Distance", y = "Standardized Speed", color = "Cluster") +
  ggtitle("Gaussian Mixture Model Clustering")
```

---

```{r cluster_dbscan, echo = FALSE}
# kNNdistplot(scaled_tracks, k = 1)
# abline(h = 2.1, col = "red")

clusters_db <- dbscan(scaled_tracks, eps = 2.1, minPts = 2)

scaled_tracks |>
  mutate(cluster = factor(clusters_db$cluster)) |>
  ggplot(aes(x = distance_km, y = speed_km_h, color = cluster)) +
  geom_point() +
  labs(x = "Standardized Distance", y = "Standardized Speed", color = "Cluster") +
  ggtitle("Density-Based Spatial Clustering")
```

---

**Ideas for more things to do?**

